Fractional Types

### 2016-12-16 typespp.tex
typos
sec 4

We call the new extension $\Pi^/$. We begin by adding four new syntactic categories to the definition of $\Pi$ in Sec.~\ref{opsempi}:

\begin{definition}[$\Pi^/$]
\label{def:pifrac}
The syntax of $\Pi^/$ is given by the following categories:
\[\begin{array}{lrcl}
(\textrm{Types}) & \tau &::=& [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{Values}) & v &::=& [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{1-combinators}) & c &:& \tau_1 \iso \tau_2 ~::= [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{2-combinators}) & \alpha &:& c_1 \isotwo c_2 \mbox{~where~}
  c_1, c_2 : \tau_1 \iso \tau_2 ~::= [\textit{see Def.~\ref{def:pi}}] \\
\\
(\textrm{Fractional Types}) & \mathbb{T} &::=&
  \order{c} \alt \iorder{c} \alt \mathbb{T}_1 \circledast \mathbb{T}_2 \\
 (\textrm{Fractional Values}) & \mathbb{V} &::=&
  c^k \alt 1/c^k \alt \langle \mathbb{V},\mathbb{V} \rangle \\
 (\textrm{/-combinators}) & \rho &:& \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::=
  [\textit{see Fig.~\ref{pifrac:comb}}] \\
(\textrm{//-combinators}) & \varrho &:& \rho_1 \leftrightblackspoon \rho_2 \mbox{~where~}
  \rho_1, \rho_2 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::=
  [\textit{see Fig.~\ref{pifrac:twocomb}}]
\end{array}\]
\end{definition}

The syntactic category $\ft$ of fractional types introduces type expressions for
iteration groupoids, symmetry groupoids, and their products. The combinators
that relate these fractional types fall into several groups. The first line
ensures that the combinators include the identity and sequential composition
$\circledcirc$ and can be applied anywhere inside a product $\circledast$. The
next two lines ensure that $\circledast$ carries a symmetric monoidal structure
with combinators witnessing unit, commutativity, and associativity. The last
line relates two types $\order{c_1}$ and $\order{c_2}$ via $\order{\alpha}$
whenever $\alpha : c_1 \isotwo c_2$. Indeed if $c_1$ and $c_2$ are considered
equivalent via a 2-combinator then we also consider their iteration groupoids to
be equivalent. The last two combinators $\eta_{\ft}$ and $\epsilon_{\ft}$ are
inspired from the definition of compact closed categories~\cite{ccc} which are
symmetric monoidal categories in which every object has a dual. In that setting,
the dual $A^*$ of an object $A$ comes equipped with two morphisms:
\begin{itemize}
\item a \emph{unit} $\eta_A : 1 \rightarrow A^* \otimes A$, and
\item a \emph{counit} $\epsilon_A : A \otimes A^* \rightarrow 1$
\end{itemize}
that satisfy appropriate coherence conditions. These conherence conditions are
added as terms in the $\varrho$ syntactic category.

\paragraph*{Values.} When the types denote sets, it is evident what it means to have a
value of a given type: it is just an element of the set. When types
denote groupoids, it is less clear what it means to have a value,
especially when the total number of values, as reported by the
groupoid cardinality, is a proper fraction. We obviously cannot list
``half a value'' but what we \emph{can} do is to list an integral
number of values and provide an equivalence relation that specifies
which values are distinguishable such that the ultimate counting of
distinguishable values is a fractional amount. The idea is not
uncommon: in the conventional $\lambda$-calculus, we list $\lambda
x.x$ and $\lambda y.y$ as separate values of type $\tau \rightarrow
\tau$ and then provide a separate equivalence relation
($\alpha$-equivalence) to express the fact that these two values are
indistinguishable. The treatment in our setting is similar but richer
as in some cases the equivalence relation is not external but is
itself part of the value and the resulting count may be fractional.

>>>>

We call the new extension $\Pi^/$. We begin by adding three new syntactic categories to the definition of $\Pi$ in Sec.~\ref{opsempi}:

\begin{definition}[$\Pi^/$]
\label{def:pifrac}
The syntax of $\Pi^/$ is given by the following categories:
\[\begin{array}{lrcl}
(\textrm{Types}) & \tau &::=& [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{Values}) & v &::=& [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{1-combinators}) & c &:& \tau_1 \iso \tau_2 ~::= [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{2-combinators}) & \alpha &:& c_1 \isotwo c_2 \mbox{~where~}
  c_1, c_2 : \tau_1 \iso \tau_2 ~::= [\textit{see Def.~\ref{def:pi}}] \\
\\
(\textrm{Fractional Types}) & \mathbb{T} &::=&
  \order{c} \alt \iorder{c} \alt \mathbb{T}_1 \circledast \mathbb{T}_2 \\
 (\textrm{Fractional Values}) & \mathbb{V} &::=&
  c^k \alt 1/c^k \alt \langle \mathbb{V},\mathbb{V} \rangle \\
 (\textrm{/-combinators}) & \rho &:& \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::=
  [\textit{see Fig.~\ref{pifrac:comb}}]
\end{array}\]
\end{definition}

The syntactic category $\ft$ of fractional types introduces type expressions for iteration groupoids, symmetry groupoids, and their products. The combinators that relate these fractional types are in Fig.~\ref{pifrac:comb}. The bottom group includes identity and sequential composition $\circledcirc$ combinators and ensures that the combinators can be applied anywhere inside a product $\circledast$. There is also a combinator that relates two types $\order{c_1}$ and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1 \isotwo c_2$. Indeed if $c_1$ and $c_2$ are considered equivalent via a 2-combinator then we also consider their iteration groupoids to be equivalent. The top group ensures that $\circledast$ carries a symmetric monoidal structure with combinators witnessing the unit, commutativity, and associativity properties. The last two pairs combinators $\eta-/\epsilon-$ and $\eta+/\epsilon+$ are inspired from the definition of compact closed categories~\cite{ccc} which are
symmetric monoidal categories in which every object has a dual. They are the ones that witness the ``fractional'' nature of symmetry groupoids: they are indeed the multiplicative inverses of iteration groupoids.

The last new syntactic category in $\Pi^/$ is that of values which deserves some background discussion. When the types denote sets, it is evident what it means to have a value of a given type: it is just an element of the set. When types denote groupoids, it is less clear what it means to have a value, especially when the total number of values, as reported by the groupoid cardinality, is a proper fraction. In that case we the values up to an equivalence relation. The idea is not uncommon: in the conventional $\lambda$-calculus, we list $\lambda x.x$ and $\lambda y.y$ as separate values of type $\tau \rightarrow \tau$ and then provide a separate equivalence relation ($\alpha$-equivalence) to express the fact that these two values are indistinguishable. The treatment in our setting is similar: for iteration groupoids $\order{c}$, every distinct iterate $c^k$ is a listed as a separate value with the understanding that some iterates are $\idisotwo$-equivalent. Similarly for symmetry groupoids $\iorder{c}$, every morphism is listed as a value $1/c^k$ with the same understanding that that some of these morphisms are $\idisotwo$-equivalent.
